---
title: "ANT 388C Lesson 3: Reproducible Research"
output: html_document
date: "2024-01-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Module 5: Basics of version control

## First steps with git and GitHub

### Installing git
Requires installing the package manager software Homebrew https://brew.sh/ in Terminal
  Requires admin privileges for erc2866, gained via https://github.com/SAP/macOS-enterprise-privileges
in Terminal
  `brew install git`
Check installation in Terminal
  `which git`
  output `/usr/local/bin/git`
  `git --version`
  output `git version 2.35.0` (mine is `git version 2.15.0`)
Install Xcode Command Line Tools
  `xcode-select --install`
  
  
### Tell git who you are
```{r}
install.packages("usethis")
library(usethis)
use_git_config(user.name = "ecarlson5683", user.email = "ecarlson5683@gmail.com")
detach(package:usethis)
```


### Generate a PAT on github.com
In the upper-right corner of any page on GitHub, click your profile photo, then click “Settings”
In the left sidebar, click “Developer settings”
In the left sidebar, click “Personal access tokens”.
Click “Generate new token”.
Give your token a descriptive name in the “Note” field. (my-github-access)
Set an expiration date for the token.
Select extent of permissions you would like to grant this token. 
  To use your token to access repositories from the command line, select the “repo” checkbox.
  Also requires read:org and workflow
Click the green “Generate token” button.
  `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`
Once you have a token, you enter it instead of your password when performing git operations over HTTPS. Personal access tokens can only be used for HTTPS git operations


### Caching Your GitHub Credentials
Requires the Homebrew package manager

#### Using the GitHub CLI (Command Line Interface)
The GitHub CLI will automatically store your git credentials for you when you choose HTTPS as your preferred protocol for git operations and answer “yes” to the prompt asking if you would like to authenticate to git with your GitHub credentials.

Install the GitHub CLI by running `brew install gh` at a command prompt in a terminal window.
In the command line, enter `gh auth login`, then follow the prompts.
When prompted for your preferred protocol for ***git*** operations, select **HTTPS**
When asked if you would like to authenticate to ***git*** with your ***GitHub*** credentials, enter “Y”


### Creating a Remote Repository
follow the instructions laid out in https://guides.github.com/activities/hello-world/ ***GitHub*** guide


## Connecting ***git*** and ***GitHub***
Making a clone of the remotely hosted repository on a local computer. We can then work on and commit changes locally via **git** and then, when we are ready, we can “push” those changes up to the remotely hosted repository on ***GitHub***.

### Cloning a Remote Repository
created a repository on github.com, then
in the RStudio terminal (Tools > Terminal > New Terminal)
  `cd /Users/erc2866/Development/Repos` (create new repos in a dedicated folder called `Repos/`)
open a web browser, go to the landing page for your repository on **GitHub.com**, click the green button that says “Code”, then select the “HTTPS” tab and copy the web URL
Return to the terminal window in RStudio or to the external shell prompt
  `git clone https://github.com/ecarlson5683/ANT388C.git`
If you then navigate into the newly created local directory using the `cd` command, you can list and look at the cloned files.  
  `cd ANT388C` (to change into the correct directory)
  `ls -a` (to list all the files in the directory)
  `head README.md` (shows the first few lines of the file “README.md”)
  `git remote show origin` (shows information about the remote repository, including the branch that “push” and “fetch” commands will be applied to and which branch is being tracked locally)


### Making and pushing local changes
edit one of the local files in your repo
  `open README.md` (opening the “README.md” file in a text editor)
Make a change to the file in the text editor or using `echo` >> <file>
  `echo "Here is some new text I am adding from the shell to update the README file." >> README.md`
  `git status` (see how file has changed)
To propagate changes back up to the remote repository: 
  [1] “stage” or “add” the files to a queue of local changes. 
  [2] “commit” those changes to our local repository so that git acknowledges that changes have been made and approved. 
  [3] “push” the committed changes up across the internet to the remote repository hosted on GitHub. 
  These are done with the git commands `add`, `commit`, and `push` in the shell:
  `add -A`
  
To propagate our changes back up to the remote repository, we now need to do three things: [1] First, we need to “stage” or “add” the files to a queue of local changes. [2] We then need to “commit” those changes to our local repository so that git acknowledges that changes have been made and approved. [3] Finally, we then need to “push” the committed changes up across the internet to the remote repository hosted on GitHub. These are done with the git commands add, commit, and push in the shell as follows:
  `git add -A`
  `git commit -m "A commit from my local repo"`
  `git push`
  
NOTE: The `-A` argument following the git add command means to stage all files in the repo that have changed since the last commit (in this case, we have only one, “README.md”). The `-m` argument following `git commit` indicates the message we want to include with our commit. It is necessary to include SOME message with each commit, and it is good practice to include a short description of what the commit includes, e.g., “Updating README file”.
  
  
### Pulling Changes from ***GitHub***
Make an edit to the README.md online, and commit it
Return to the shell prompt inside the local directory for your repository and enter the command 
  `git pull`
You should see something indicating that the main branch of the local repository has been “fast-forwarded” to the state of the remote main branch that is being tracked:
If you run the command `git pull` and the local repo is in sync with the version hosted on GitHub, you should see something like the following:
  Already up to date.


## Additional useful information
See Module 5.6 for "Using the Shell" and SSH file transfer protocol (instead of HTTPS) 
with `RSA` key pairs


###Reproducible research

Introduction to GitHub, R projects, R Markdown, and Quarto


Add a repository to github user interface

github.com/ecarlson5683
Profile > Your repositories > New
Title: ANT388C
Select Public (if you want other people to see it)
Select "Add a README file"
Add .gitignore > template for R
  tell git not to watch changes to R data file as a default
Create repository


To clone this repository down to local RStudio
Code tab > https://github.com/ecarlson5683/ANT388C.git

in RStudio > File > New Project > Create from Version > Git
then paste in git link https://github.com/ecarlson5683/ANT388C.git
Can choose where you want to create repository on local machine
Create project

Should prompt to sign into github account

How to add a collaborator
in github > "<Repository of interest>" >Settings > Collaborators


Create a new RMarkdown document
in RStudio > File > New File > RMarkdown >
  document box: Title, etc.
  opens a template for R Markdown file
  
Create a new Quarto Document
in RStudio > File > New File > RMarkdown >
  document box: Title, etc.
  opens a template for Quarto file
  
To put a flog, write a Commit message

```{r}

```

###Module 6:


###Exercise 02 (due before class on Monday)

Set up GitHub repo in GitHub workspace
Add Dr. Di Fiore to repo (username "difiore")
Clone repo to your computer
Create an RMarkdown or Quarto document
