---
title: "ANT388C_12_central_tendency_and_spread"
author: "Erika Carlson"
format: html
editor_options: 
  chunk_output_type: console
---

# Class 8: Functions and the Wordle Coding Challenge

## Functions

Function psuedo code

```{r}
my_function <- function(<argument list>) {
  <function code>
  return(<value>)
}
```
functions should have certain default values than can be modified

A toy function
```{r}
# x should be tabular data
my_function <- function(x, reps = 2){
  for (i in 1:reps){
    print(x)
  }
  for (i in 1:nrow(x)){
    print(x[i, ])
  }
  return(NULL)
}

df <- tibble(name = c("Erika", "Ryan", "Steven", "Stephen"), 
            year = c(3, 6, 6, 3),
            hometown = c("Elburn", "Bardstown", "Albuquerque", "El Paso"))

my_function(df)
```


a different toy function
```{r}
# x should be tabular data
my_function <- function(x, condition, variable){
  library(tidyverse)
  x <- x %>% filter(variable %in% condition)
  return(x)
}
```


## Inner Joins

Matching items in the left- and right-hand tables, only gives index matches

## Outer Joins

* `left_join()` pulls out unique joins from the left-hand table, will fill in `NA` for unmatched right-hand rows
* `right_join()` pulls out unique joins from right-hand table, will fill in `NA` for unmatched left-hand rows
* `full_join()` pulls out unique joins from both left- and right-hand tables, will fill in `NA` for unmatched rows

## Practicing Joins

Load 2 datasets from GitHub

  **papers.csv as "p"**
  
  **creators.csv as "c"**
  
```{r}
c <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/creators.csv"
c <- read_csv(c, col_names=TRUE)

p <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/papers.csv"
p <- read_csv(p, col_names=TRUE)

# split author list
library(tidyverse)
p <- p %>% 
  separate_wider_delim(cols = Author,
                       delim = ";",
                       names = c("First Author", "A2", "A3", "A4"),
                       too_few = "align_start", too_many = "drop")
head(p)

c <- c %>% 
  distinct()
head(c)
```

Join data frames
```{r}
# these give same number of rows, but in different order and diff index header
inner <- inner_join(c, p, by = c("fullName" = "First Author"))
inner <- inner_join(p, c, by = c("First Author" = "fullName"))

# left join
left <- left_join(c, p, by = c("fullName" = "First Author"))

# right join
right <- right_join(c, p, by = c("fullName" = "First Author"))
```

Find publications
```{r}
find_pubs <- tibble(fullName = c("Di Fiore, Anthony"))
first <- inner_join(find_pubs, p, by = c("fullName" = "First Author"))
view(first)
```



